Red Wine Quality Exploratory Data Analysis by Saul Geller
========================================================



```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
knitr::opts_chunk$set( fig.width=9, fight.height=5, fig.path='Figs/',
                       fig.align='center', tidy=TRUE,
                       echo=FALSE, warning=FALSE, message=FALSE)

# Here I add the libraries I will be using

library(ggplot2)
library(dplyr)
library(knitr)

```

```{r echo=FALSE, message=FALSE, warning=FALSE, Load_the_Data}
# Here I load the data into a csv, and check for missing and null values.
# Then I rename a few columns to my preferences and print out my check on missing and null values.

rw_df <- read.csv("wine_qual_reds.csv")
missing.values <- sum(is.na(rw_df))
null.values <- sum(is.null(rw_df))
rw_df <- rename(rw_df, sulfates = sulphates, fixed.acid = fixed.acidity, 
                vol.acid = volatile.acidity, res.sugar = residual.sugar, 
                free.sulfur = free.sulfur.dioxide, total.sulfur = total.sulfur.dioxide)
cat("Number of missing values:", missing.values)
cat("\n")
cat("Number of null values:", null.values)

```


The dataframe contains around 1600 entries for various red wines. Each entry contains 
a quality rating score between 0 and 10, and 11 chemical properties, all with numerical 
values. There are no missing or NULL values so no data cleaning is necessary at this 
point. I did decide to change the column named "sulphates" to "sulfates" just to
be more consistent with the spelling, and I also shortened some column names.



# Univariate Plots Section



```{r echo=FALSE, Univariate_Plots}
#The standard summary function gives me a good view of the dataset.

summary(rw_df)


```


Looking at the summary of the columns in the dataframe, nothing crazy stands out. There do 
seem to be some relatively high maximum values compared to the minimum and median values, indicating 
some potential outliers. The pH values also look surprisingly acidic, though I don't know much about 
wine pH going into this project.

I also want to make a note of the units of the data, given by the data info text document provided 
by the authors of the dataset. Most of these columns are given in some form of density unit; the only 
exceptions being pH, percent alcohol, and the quality score. The unit g/dm^3, which applies to fixed 
acidity, volatile acidity, citric acid, residual sugar, chlorides, and sulphates, means grams per 
decimeter cubed. A decimeter cubed is also 1 liter, so more simply this unit is grams per liter.

The free sulfur dioxide and total sulfur dioxide are given in mg/dm^3 which is just milligrams per 
liter, and the overall density of the wine is given in g/cm^3 which is also grams per milliliter.

```{r echo=FALSE}
# To create histogram plots for each column in the dataset, I use ggplot and the geom_histogram function.
# Using the scale_x_continuous function I can also control where the x axis is labeled.
# Here I create a histogram of the fixed.acid distribution.

ggplot(aes(x=fixed.acid), data=rw_df) + geom_histogram(bins=60) + 
  scale_x_continuous(breaks=seq(4,16,1))

```


```{r echo=FALSE}
# Here I do the same thing for the vol.acid column.

ggplot(aes(x=vol.acid), data=rw_df) + geom_histogram(bins=60) + 
  scale_x_continuous(breaks=seq(0,1.6,0.2))

```


The fixed and volatile acidity level distributions both look fairly normal, but with some 
right-skewedness. Both these columns measure the density of different kinds of acid in the 
wine. Fixed acidity measures tartaric acidwhile volatile acidity measures acetic acid. 
High levels of acetic acid can cause a vinegar taste in the wine, as noted in the author 
provided data info text file.

```{r echo=FALSE}
# Here is the citric.acid distribution.

ggplot(aes(x=citric.acid), data=rw_df) + geom_histogram(bins=50) + 
  scale_x_continuous(breaks=seq(0,1,0.1))

```


The citric acid distribution shows an unusually high number of wines at 0 and near 
0.5 g/dm^3. Overall thedistribution does not appear very normal but other than those 
2 irregular peaks, it looks like the count of wines with citric acid levels from 0 to 0.5 
fluctuates and then starts falling off at higher citric acid levels. There is also an outlier 
near 1.

```{r echo=FALSE}
# Here is the res.sugar distribution.

ggplot(aes(x=res.sugar), data=rw_df) + geom_histogram(bins=50) +
  scale_x_continuous(breaks=seq(0,16,1))

```



```{r echo=FALSE}
# The previous distribution was very long-tailed so using a scale transformation I can change that.
# In this case I used a scale_x_log10 function to make the x scale logarithmically.

ggplot(aes(x=res.sugar), data=rw_df) + geom_histogram(bins=50) + 
  scale_x_log10(breaks=c(1,2,3,5,7,9,11,13,15))

```


The original residual sugar distribution looks long tailed so I applied a logarithmic 
scale transformation. The transformed distribution appears much more normal than before, 
but is still somewhat long-tailed.

```{r echo=FALSE}
# Here is the chlorides column distribution.

ggplot(aes(chlorides), data=rw_df) + geom_histogram(binwidth=0.01) + 
  scale_x_continuous(breaks=seq(0,0.7,0.1))

```

```{r echo=FALSE}
# Again the previous distribution is long tailed so I want to try a scale transformation.

ggplot(aes(chlorides), data=rw_df) + geom_histogram(binwidth=0.01) + 
  scale_x_log10(breaks=c(0.01,0.02,0.03,0.04,0.06,0.08,0.10,0.15,0.2,0.3,0.4,0.6))

```


The chlorides level is basically the density of various salts in the wine. The original 
distribution looked long-tailed and applying a logarithmic scale transformation the new 
distribution looks much more normal, with a mean value at around 0.9.

```{r echo=FALSE}
# Here is the distribution for free.sulfur.

ggplot(aes(x=free.sulfur), data=rw_df) + geom_histogram(bins=40) + 
  scale_x_continuous(breaks=seq(0,70,5))
```

```{r echo=FALSE}
# Again the previous distribution is long-tailed.
# This time I used the scale_x_sqrt function to scale the x axis.

ggplot(aes(x=free.sulfur), data=rw_df) + geom_histogram(bins=40) + 
  scale_x_sqrt(breaks=c(1,2,3,4,6,8,10,15,20,25,30,40,50,60,80))

```

```{r echo=FALSE}
# Here is the total.sulfur distribution.

ggplot(aes(x=total.sulfur), data=rw_df) + geom_histogram(bins=40) + 
  scale_x_continuous(breaks=seq(0,300,25))

```

```{r echo=FALSE}
# Here is the total.sulfur distribution with a logarithmic scale distribution.

ggplot(aes(x=total.sulfur), data=rw_df) + geom_histogram(bins=40) + 
  scale_x_log10(breaks=c(5,10,25,50,75,100,150))

```


Sulfur dioxide (SO2) is a biproduct of fermentation, but most winemakers actually add 
significantly more of it as a preservative and anti-microbial agent. The SO2 can either be 
free, where it exists in equilibrium between its molecular form as well as two ionic forms 
in solution, or it can be bound, meaning its chemically attached to another compound in the 
wine. The total SO2 column is the combination of both free and bound SO2. Both SO2 
distributions appear long-tailed, so I applied scale transformations to get them to look 
more normal. 

For the Free SO2 distribution, a square root transformation worked best. There is an 
unusually large peak at around 5-6 but other than that the new distribution appears normal. 

For the Total SO2 distribution, a logarithmic scale transformation worked better and overall 
it appears quite normal in the new graph, with its mean at around 46.5.



```{r echo=FALSE}
# Here is the sulfates column distribution.

ggplot(aes(x=sulfates), data=rw_df) + geom_histogram(binwidth=0.02) + 
  scale_x_continuous(breaks=seq(0,2,0.25))

```


```{r echo=FALSE}
# Here is the sulfates distribution with a logarithmic scale transformation.

ggplot(aes(x=sulfates), data=rw_df) + geom_histogram(binwidth=0.02) + 
  scale_x_log10(breaks=c(0.3,0.4,0.5,0.6,0.7,0.8,1.0,1.2,1.4,1.6,1.8,2.0))

```


Sulfates are wine-additives that can add to the sulfur dioxide levels of the wine. The 
original distribution looked somewhat long-tailed and a logarithmic scale transformation
makes the new graph look much more normal, with a mean of around 0.66.

```{r echo=FALSE}
# Here is the density distribution.

ggplot(aes(x=density), data=rw_df) + geom_histogram(bins=40) + 
  scale_x_continuous(breaks=seq(0.990,1.005,0.0025))

```


The wine density distribution looks quite normal, with all of its values very close to 1, 
the density of water. the distribution is not centered at 1 however. Most of the wines have 
a density slightly less than 1, and the overall mean is 0.997.

```{r echo=FALSE}
# Here is the pH distribution.

ggplot(aes(x=pH), data=rw_df) + geom_histogram(bins=40) + 
  scale_x_continuous(breaks=seq(2,5,0.1))

```


The pH is a measure of acidity based on the concentration of hydrogen cations in a 
solution. The typical scale goes from 0 to 14, with a pH of 7 being the pH of water and 
considered neutral. Anything less than 7 is acidic, and anything more than 7 is basic. 

For this dataset, the pH distribution appears normal, centered around 3.3. All of the 
wines are below 7.

```{r echo=FALSE}
# Here is the percent alcohol distribution of the wines.

ggplot(aes(x=alcohol), data=rw_df) + geom_histogram(bins=40) + 
  scale_x_continuous(breaks=seq(8,15,1))

```


The percent alcohol distrubtion does not look normal. It is skewed right, with a large 
peak around 9.5, and then the counts decrease as the percent alcohol increases. Applying 
square root or logarithmic scale transformations does not normalize this distribution.

```{r echo=FALSE}
# Here is the distribution of the quality scores.

ggplot(aes(x=quality), data=rw_df) + geom_histogram(binwidth=1) + 
  scale_x_continuous(breaks=seq(1,10,1), minor_breaks=NULL)
```


Quality scores go from 0 to 10, but in this dataset, only scores from 3 to 8 actually 
occurred. The most common scores were 5 and 6, and the overall distribution looks normal. 
The mean quality score is 5.6.



# Univariate Analysis



### What is the structure of your dataset?

The dataset is made up of 13 columns, with 1599 observations. The first column is just the
unique number given to each wine in sequence, going from 1 to 1599, so really there are only
12 variables. Of those 12 variables, 11 are measures of various chemical properties, mainly
densities of particular compounds in the wine. The 12th variable is the quality score of 
the wine. 

All 12 variables are numerical, with no real ordered factor variables. That said, the quality 
column can be treated like an ordered factor variable because it is an integer score that can
only go from 0 to 10. And in this dataset, it only goes from 3 to 8.


### What is/are the main feature(s) of interest in your dataset?

The main feature of this dataset is the quality score, which goes from 0 to 10 with a higher
score signifying a better rated wine.


### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

Besides the quality score, the other features of the dataset are all chemical properties of 
the wine and may be linked to higher or lower quality wine. 

### Did you create any new variables from existing variables in the dataset?

I did not add any columns to the data. I thought about adding a column for bound sulfur dioxide,
which would be the total sulfur dioxide subtracted by the free sulfur dioxide, but after
investigating why sulfur dioxide levels are measured, plus the total sulfur dioxide will of course
be related to the bound sulfur dioxide, I decided against it.


### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The dataset did contain some unusual distributions, but most were fairly normal, and some became 
normal after a scale transformation. Using logarithmic or square root scale transformations on any 
distributions that appeared long-tailed typically helped to normalize them, except for the alcohol
content distribution.



# Bivariate Plots Section



```{r echo=FALSE, Bivariate_Plots}
# Here I make a correlation matrix using the cor function on the entire dataframe, minus the first column.
# Also applying the round function keeps the correlations rounded to 2 decimal places.

round(cor(rw_df[,2:13]), 2)

```


Here is a correlation matrix of all columns in the dataset, excluding the first column which
is just a sequential list of the entries. Looking at the main output in the quality column, 
the chemical property with the strongest linear relationship is the alcohol content with a
correlation of 0.48. The quality score's relationship with volatile acid is also relatively
strong. And the weakest relationships are with residual sugar, free sulfur dioxide, and pH.

Looking at other variables, I see density has a relatively strong positive linear relationship 
with fixed acidity level and an inverse linear relationship with alcohol content. Fixed acidity 
level also has a direct relationship with citric acid content, which makes sense as citric acid
levels should directly contribute to fixed acid levels since citric acid is a fixed acid.

```{r echo=FALSE}
# To graph a scatter plot I would normally use geom_point, but since the quality column uses only 
# integer values, I'll use geom_jitter so the points aren't all on the same lines. 
# Also using the alpha setting in geom_jitter can help combat overplotting.

ggplot(aes(x=alcohol, y=quality), data=rw_df) + geom_jitter(alpha=0.5, width=0.025, height=0.25)

```


```{r echo=FALSE}
# Using the facet_wrap function, I can treat quality like an ordered factor and graph multiple
# boxplots with geom_boxplot over all the different quality scores to visualize each distribution.

ggplot(aes(y=alcohol), data=rw_df) + geom_boxplot() + facet_wrap(~quality)

```


Alcohol content and quality score had a correlation of 0.48. The scatterplot shows that
wines with scores of 8 and 7 typically have more alcohol than wines with a score of 3 or
4.

Using boxplots gives a better picture of the data. From this I see that as wine quality
score goes up from 6 to 7 to 8, alcohol content also increases for the bulk of the wines.
However, ignoring the many outliers present in wines with a score of 5, we can see the bulk 
of those wines actually had less alcohol content than for wines with a score of 4.

```{r echo=FALSE}
# I use geom_jitter again to add some noise to the y axis of this scatter plot.

ggplot(aes(x=vol.acid, y=quality), data=rw_df) + geom_jitter(alpha=0.5, width=0, height=0.25)
```


```{r echo=FALSE}

# Here are the distributions of vol.acid over the different quality scores using geom_boxplot
# and facet_wrap. 

ggplot(aes(y=vol.acid), data=rw_df) + geom_boxplot() + facet_wrap(~quality)
```


Again, the relationship in the scatter plot is apparent, though this time its an inverse
relationship. Looking at the boxplots, wines with a score of 8 and 7 are similar, but
as the wines' score gets lower, volatile acid level increases in a pretty linear fashion.

Volatile acid level is basically the amount of acetic acid in the wine, which at high levels
will cause a vinegary taste.


```{r echo=FALSE}
# Again I use geom_jitter to add some noise and help prevent overplotting when looking at
# quality vs citric.acid in this scatter plot.

ggplot(aes(x=citric.acid, y=quality), data=rw_df) + geom_jitter(alpha=0.5, width=0, height=0.25)
```


```{r echo=FALSE}
# Using geom_boxplot and facet_wrap I make another visualization, this time of citric.acid's
# distributions over the different quality scores.

ggplot(aes(y=citric.acid), data=rw_df) + geom_boxplot() + facet_wrap(~quality)

```

According to the authors of the dataset, "citric acid can add freshness and flavor to wines."
Looking at citric acid content graphed against quality score, wines with scores of 7 and 8
do seem to have higher amounts of citric acid, but that pattern doesn't really hold true for
lower scores. The overall relationship is also quite weak with a correlation of 0.23.

Looking specifically at wines with a score of 7 vs wines with a score of 6, the 6 score wines
are pretty evenly spread in terms of of citric acid level, but 7 score wines have a clear lack
of data points around 0.2 g/dm^3 citric acid, with a grouping of outlier points that are have
lower citric acid levels.


```{r echo=FALSE}
# Using geom_point I noticed too many discrete lines so switching to jitter to add some random
# noise to the free.sulfur datapoints makes these lines disappear.

ggplot(aes(x=free.sulfur, y=pH), data=rw_df) + geom_jitter(alpha=0.5, width=0.25, height=0)
```

```{r echo=FALSE}
# Here is sulfates vs free.sulfur using geom_jitter to get rid of some of the discrete lines
# made by free.sulfur's values.

ggplot(aes(x=sulfates, y=free.sulfur), data=rw_df) + geom_jitter(alpha=0.5, width=0, height=0.25)
```


One thing I was curious about was how free sulfur dioxide levels related to some of the
other variables. I looked at free sulfur versus pH as well as free sulfur vs sulfates.

Free sulfur exists in a chemical equilibrium between multiple states. It will exist as a
gas, molecular sulfur dioxide, as a bisulfite ion with a negative charge, and as a sulfite 
ion with two negative charges, by giving up a hydrogen ion. This relationship made me
wonder if free sulfur dioxide levels could have a link to pH. According to the correlation 
matrix there was no linear relationship but I still wanted a closer look.

Sulfates is a measure of how much sulfate preservatives were added to the wine. The authors
say that these additives should add to free sulfur levels but according to the correlation matrix
there was no relationship between them.



```{r echo=FALSE}
# There was some overplotting in this scatter plot so I set geom_point's alpha setting to 0.5.

ggplot(aes(x=fixed.acid, y=density), data=rw_df) + geom_point(alpha=0.5)
```


```{r echo=FALSE}
# Here is alcohol vs density with some jitter added for the alcohol column.

ggplot(aes(x=alcohol, y=density), data=rw_df) + geom_jitter(alpha=0.5, width=0.025, height=0)
```

```{r echo=FALSE}
# Here is fixed acid vs alcohol with some jitter added to both dimensions.

ggplot(aes(x=fixed.acid, y=alcohol), data=rw_df) + geom_jitter(alpha=0.5, width=0.25, height=0.025) 
```


The density of the wines was moderately correlated to fixed acidity level and alcohol content.
Alcohol, specifically ethanol, has a lower density than water, so a wine with higher 
alcohol content having a lower density makes sense. But, I wasn't sure how fixed acidity
level relates. Looking at the fixed acidity vs the density, the direct linear relationship
is very noticeable.

Comparing fixed acidity to alcohol content, there's no real pattern that I can see, and the
correlation is also quite weak which leads me to believe that however fixed acidity level and
density are linked, its independent from alcohol content's effect on density.



# Bivariate Analysis



### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The main feature of this dataset, the quality score, correlated most strongly
with alcohol content and volatile acid level. Higher alcohol levels were associated
with better quality scores and higher volatile acid levels were associated with
lower quality scores. Percent alcohol having this strong an effect on quality
(with a correlation of 0.48) surprised me, as I don't think a wine needs to be
very strong to be good. The quality's relationship with volatile acidity is far
less surprising as according to the authors, a high volatile acidity level means
a vinegary taste.

I was also surprised by the lack of relationship between quality and citric acid
level as according to the author, citric acid adds flavor and freshness to a wine.
And there was no real relationship between pH and quality which also surprised me,
as I thought pH could definitely effect taste. Of course there are many drinks have
even lower pH's, such as coca cola soft drinks have a pH of around 2.5.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

Density had a surprisingly strong relationship with fixed acidity level, and
free sulfur dioxide level had a surprisingly weak to no relationship with both
pH and sulfates added.

### What was the strongest relationship you found?

Fixed acidity level had the strongest relationships in the dataset, with a 
correlation of 0.67 to citric acid level, 0.67 to density, and -0.68 to pH. 
Most of these relationships are not suprising. Citric acid is a type of fixed
acid so of course they are related. And having a lot of fixed acid will of course
lower the pH. However, as stated previously, density having a moderately strong
correlation with fixed acidity level was unanticipated.



# Multivariate Plots Section



```{r echo=FALSE}
# Using facet_wrap I can separate the scatter plot distributions by quality.

ggplot(aes(x=free.sulfur, y=pH), data=rw_df) + geom_point(alpha=1/3) + facet_wrap(~quality)
```


Here is a depiction of free sulfur dioxide vs pH over different qualities. The 
distributions are all pretty similar. Wines with an 8 quality score look somewhat
positively correlated but with too few points its probably not statistically
significant.




```{r echo=FALSE}
# Using fixed.acid as a variable to determine the color of a point, I can add another level
# of depth to this visualization.

ggplot(aes(x=free.sulfur, y=pH), data=rw_df) + geom_point(aes(color=fixed.acid)) + 
  facet_wrap(~quality)
```

Here is free sulfur vs pH over different qualities, with a color gradient representing
the fixed acidity level. Higher fixed acidity levels causes a lower pH, and I wanted to 
see if maybe that effect was masking any effects free sulfur dioxide could have on the pH.

Darker colors are more common at higher levels of pH, which was already shown in the 
bivariate section looking at the relationship between fixed acidity and pH. I don't see 
anything that indicated fixed acidity levels are masking free sulfur's effect on pH, and
this holds true for all qualities of wine.


```{r echo=FALSE}
# Using color as a variable again, here is a scatter plot of fixed.acid vs density vs alcohol.

ggplot(aes(x=fixed.acid, y=density), data=rw_df) + geom_point(aes(color=alcohol))

```



```{r echo=FALSE}
# Using color and facet_wrap to look at fixed.acid vs density vs alcohol over the different
# quality scores.
ggplot(aes(x=fixed.acid, y=density), data=rw_df) + geom_point(aes(color=alcohol)) +
  facet_wrap(~quality)

```

I then looked at the interesting relationship between density and fixed acidity. 
Density is negatively associated with alcohol content because alcohol is less dense
than water, but I wasn't sure why fixed acidity had any relationship with density.

Looking at fixed acidity vs density, with a color gradient representing alcohol content,
as well as looking at it over different qualities of wine, there looks to be a very strong
relationship between high percent alcohol wines and fixed acidity, though they have no
correlation when I looked at just fixed acidity vs alcohol in the bivariate section.

Though the most obvious relationship is between alcohol and density, most of the ligher
color points are on the left side of the first graph. Looking at this split into
different graphs for every quality, this effect holds true mainly only for wines with scores
of 4, 5, and 6. While wines with scores of 3 and 8 had too few points to establish a pattern, 
wines with a quality of 7 seemed to have less of a relationship between alcohol and fixed
acidity.

```{r echo=FALSE}
# Here is free.sulfur vs total.sulfur vs sulfates. I used geom_jitter again because
# the free.sulfur and total.sulfur columns have only integer values and some discrete lines 
# were noticeable.

ggplot(aes(x=free.sulfur, y=total.sulfur), data=subset(
  rw_df, sulfates < quantile(rw_df$sulfates, 0.95))) + 
  geom_jitter(aes(color=sulfates), width=0.25, height=0.25) 
```

I also looked at sulfates vs free sulfur dioxide in the bivariate section but I wanted to 
try seeing if there was a pattern when I included total sulfur dioxide as well, as 
perhaps the sulfur additives were getting bound and would thus not be free. 

Because the sulfates distribution had some far right outliers, I only used data up to the
95th percentile of the sulfates distribution so the color differences would be more even.

However, there didn't seem to be any apparent pattern to sulfate level versus either of
these variables. There is a clear positive association between free sulfur and total sulfur, 
which is to be expected.

```{r}
# Here I use R's built in lm function to create a linear model from the dataset. 
# Then the summary function gives a good summary of that linear model.
# Lastly I find the mean of the residuals of the linear model because one of the assumptions
# of linear regression is that the mean of the residuals is 0.

alc.qual.lm <- lm(formula = quality ~ alcohol, data=rw_df)
summary(alc.qual.lm)
mean(alc.qual.lm$residuals)
```

```{r}
# Here is a similar linear regression model but adding in the other factors that had the highest
# correlations with quality.

qual.multiple.lm <- lm(quality ~ alcohol + vol.acid + citric.acid + sulfates, data=rw_df)
summary(qual.multiple.lm)
mean(qual.multiple.lm$residuals)
```


I used linear regression to try and find a model to best explain the quality
score output based on certain chemical properties of the wine. Looking at just 
alcohol as the explanatory variable there was a clear statistically significant
effect on quality. With an R-squared value of 0.226, alcohol level only 
explains about 23% of the quality scores variance.

Adding in other variables that shared a correlation with quality, and none that
were associated with each other, the R-squared value improved to about 0.334. One
of the variables, citric acid level, did not have a statistically significant
relationship however.

```{r}
# Here is another linear regression model looking at density instead of quality, with 
# fixed.acid as the explanatory variable.

density.lm <- lm(density ~ fixed.acid, data=rw_df)
summary(density.lm)
mean(density.lm$residuals)
```

```{r}
# And here is another model for density but adding in alcohol as a contributing factor.

density.multiple.lm <- lm(density ~ fixed.acid + alcohol, data=rw_df)
summary(density.multiple.lm)
mean(density.multiple.lm$residuals)
```


Even though density is not my main output of this dataset, I was still interested
in how and why fixed acidity had a positive correlation with it. So using linear
regression, fixed acidity does have a statistically significant relationship
with density and by itself has an R-squared of 0.45. Combined with alcohol, which 
had no correlation with fixed acidity, the R-squared jumps to 0.65.


# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

In my investigation, though I thought free sulfur levels might have an effect
on pH, I did not find any evidence of that in the bivariate section. Looking at 
it across different qualities, I still did not find any pattern or associations.
So as far as I can tell, free sulfur did not have an effect on pH, despite how
free sulfur dioxide in the form of bisulfites in solution can dissassociate further
and add more hydrogen cations to solution.


### Were there any interesting or surprising interactions between features?

Looking at fixed acidity versus density, a direct correlation became apparent
in the bivariate section that I had no explanation for. Looking at it compared 
with percent alcohol levels, there did seem to be a hidden relationship between
alcohol content and fixed acidity that only held true for certain quality scores.
Higher alcohol content causes lower density which is why I thought alcohol content
might also contribute to how fixed acidity relates to density.

Without further investigation I can't say if this relationship between alcohol
and fixed acidity that only occurred at certain qualities is causal or even real;
it may just be statistical noise. But it may be something interesting to look into
with further data.

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I did create some standard linear regression models to try and fit a line
that would explain how quality score varied with other chemical properties of 
the wine. The model did find clear statistical evidence that alcohol, volatile 
acidity, and sulfates had an effect on quality, but the linear model only 
explained about 33% of how quality varied. 

I also looked at how density could be explained through fixed acidity and
alcohol content. 

------

# Final Plots and Summary



### Plot One
```{r echo=FALSE, Plot_One}
# Here I use ggplot and geom_histogram again to create a single variable distribution.
# Using the labs function I can also easily add a title and labels where needed.

ggplot(aes(x=total.sulfur), data=rw_df) + geom_histogram(bins=40, color="black", fill="red") + 
  scale_x_log10(breaks=c(5,10,25,50,75,100,150)) +
  labs(title="Total Sulfur Dioxide Distribution", x="Total Sulfur Dioxide (mg/dm^3)")

```

### Description One

Here is the total sulfur dioxide distribution with a logarithmic scale transformation. 
On a linear scale, the distribution was long-tailed, but with this logarithmic scale the
distribution looks very normal, albeit with some significant variance. This tells me that
most wines have a relatively low total sulfur dioxide level, yet there are then many
outliers who have much higher sulfur dioxide levels compared to the bulk of the wines.

### Plot Two
```{r echo=FALSE, Plot_Two}
# I recreate another one of my visualizations but use ggtitle and ylab for the labeling
# instead of the labs function. I also set breaks=NULL in the scale_x_continuous function
# because there are no need for breaks since the x axis isn't measuring anything.

ggplot(aes(y=alcohol), data=rw_df) + geom_boxplot() + facet_wrap(~quality) +
  scale_x_continuous(breaks=NULL) +
  ggtitle("% Alcohol by Quality Score") +
  ylab("% Alcohol")

```

### Description Two

In the correlation matrix I saw that alcohol content had the strongest linear relationship
to the main output of the dataset, quality score. Looking at the boxplots of % alcohol 
separated by quality, I can see that wines with a score of 8 had a higher alcohol content
than wines with a score of 7, which were higher in alcohol than wines with a score of 6,
which were higher than wines with lower scores. But, interestingly, wines with a score of 
3, 4, and 5 did not have very different alcohol levels, though the wines with a score of 5's
distribution did have a lot of outliers. 

### Plot Three
```{r echo=FALSE, Plot_Three}
# In this plot I recreate a visualization from a previous section, then use ggtitle, xlab, ylab,
# and labs to label everything I want labelled. I could also only use the labs function but I
# like having everything in separate layers.

ggplot(aes(x=fixed.acid, y=density), data=rw_df) + geom_point(aes(color=alcohol)) +
  facet_wrap(~quality) +
  ggtitle("Fixed Acidity vs Density and Alcohol Content") + 
  xlab("Fixed Acidity Level (g/dm^3)") +
  ylab("Density (g/cm^3)") +
  labs(color="% Alcohol")

```

### Description Three

The most interesting relationship I found was between fixed acidity and density. 
They had a correlation of 0.67, and a positive association is apparent looking at
this graph. Alcohol content was also associated with density in an inverse relationship.
That relationship was to be expected since alcohol has a lower density than water, and
wine has a very close density to water. Looking at how percent alcohol could be 
contributing to any relationship between fixed acidity and density, I graphed all
three variables together, plus split by quality score. And despite alcohol content having 
no correlation to fixed acidity level, it did appear that alcohol was typically lower 
when fixed acidity was lower for wines with a quality score of 6 and 7. Those graphs appear
lighter on the right side and darker on the left side. However that's actually the opposite
of what would be true if percent alcohol was a contributor to fixed acidity's relationship 
to density, since percent alcohol is negatively associated with density, I would expect
it to be negatively associated with fixed acidity. This hidden relationship between 
percent alcohol and fixed acidity may just be statistical noise.

------

# Reflection


Looking through this dataset, I was able to determine that alcohol content, 
volatile acidity, and sulfate additive level were all statistically significant
contributors to quality score. Sulfates are added to wine to act as antioxidants
anti-microbial agents, and their level had a positive association with quality.
Volatile acidity is basically the amount of acetic acid present in the wine,
which at higher levels can give the wine a vinegar taste, and had a negative
association with quality. Alcohol content had a positive relationship with 
quality, but only up to a point, as lower scoring wines had pretty similar
alcohol content distributions.

There were many suprising things I found in the dataset, such as free sulfur
dioxide's lack of a relationship with pH, alcohol's relatively strong relationship
with quality, as well as fixed acidity's relationship with density. I struggled
to find an explanation for why fixed acidity had a somewhat strong (correlation of
0.67) relationship with density. Looking at how percent alcohol could maybe 
contribute to it didn't help unfortunately.

For future work I would want to look into this relationship between density and fixed
acidity, and into fixed acidity in general to establish how it can affect density.
Now, density in the context of this dataset, isn't a very important as it doesn't
have much of a relationship with the main output, quality score. But I am still
curious as to how fixed acidity would affect density, and finding out why could
reveal interesting findings, even if they wouldn't be very crucial to the context
of this particular dataset.

Also I would probably look for additional chemical properties to test against quality 
score to see if maybe a better model can be found to predict quality.
